# Classification and Survival Analysis

```{r include=FALSE}
library(knitr)
library(caret)
library(mlbench)
library(glmnet)
library(tidyverse)
library(e1071)
library(MASS)
library(party)
library(pec)
library(tidyverse)
```

We so far focused on the linear regression model. We will now extend to the setting where the response is binary or time-to-event. The former is also referred to as classification.

## Classification

We focus on the binary situation where response $G$ takes values $0$ and $1$, and, the aim is to do prediction based based on covariates $X=(X_1,\ldots,X_p)$. We model the probability of success $p(x;\beta)=P(G=1|X=x;\beta)$ using the logistic regression approach with logit link function

\[\rm{logit}(x;\beta)=log \Big(\frac{p(x;\beta)}{1-p(x;\beta)}\Big)=X^T\beta.\]

In logistic regression we estimate the regression parameter $\beta$ by maximizing the log-likelihood

\begin{align*}
\ell(\beta|\bf G,\bf X)&=\sum_{i=1}^{n}(1-g_i)\log(1-p(x_i;\beta))+g_i\log p(x_i;\beta)\\
&=\sum_{i=1}^{n}g_i x_i^T\beta - \log(1+\exp(x_i^T\beta)).
\end{align*}

Given an estimate $\hat \beta$ (from training data), prediction based on new input data $X_{\rm new}$ can be obtained via the predicted probablity of success $p(X_{\rm new};\hat \beta)$, e.g. the class labels corresponding to the maximum probability

\begin{align*}
  \hat{G}(X_{\rm new})=\left\{
    \begin{array}{ll}
      1, & \mbox{if $p(X_{\rm new};\hat\beta)>0.5$}.\\
      0, & \mbox{otherwise}.
    \end{array}
  \right.
\end{align*}

Similar as in linear regression, in the high-dimensional context where $n$ is small compared to $p$, the maximum likelihood estimator does often lead to overfitting and a poor generalisation error. We will in the following generalize the previously introduced idea of regularization to classification. In the context of linear regression we penalized the residual sum of squares (RSS). We extend this and penalize the negative log-likelihood with an elastic net penalty

\begin{align*}
\hat{\beta}_{\alpha,\lambda}&=\rm{argmin}-\frac{1}{n}\ell(\beta|\bf G,\bf X)+\lambda\big((1-\alpha)\|\beta\|_2^2/2+\alpha\|\beta\|_1\big).
\end{align*}

There are different measures to judge the quality of the predictions. We focus on the missclassification error  which is simply the fraction of missclassified test samples. Another important measure used in the context of classification is the receiver operating characteristic (ROC). 

### South African Heart Disease Data

We start with a "low" dimensional example illustrating the logistic regression model and comparing 
the Lasso with traditional backward selection. The data in Figure \@ref(fig:sahd) are a
subset of the Coronary Risk-Factor Study (CORIS) baseline survey, carried
out in three rural areas of the Western Cape, South Africa (Rousseauw et al, 1983, South African Medical
Journal). The aim of the study was to establish the intensity of ischemic
heart disease risk factors in that high-incidence region. The data represent
white males between 15 and 64, and the response variable is the presence or
absence of myocardial infarction (MI) at the time of the survey (the overall
prevalence of MI was 5.1% in this region). There are 160 cases in our data
set, and a sample of 302 controls. This example is taken from the book "Elements of Statistical Learning". The variables are:

- sbp: systolic blood pressure
- tobacco: cumulative tobacco (kg)
- ldl: low densiity lipoprotein cholesterol adiposity
- famhist: family history of heart disease (Present, Absent)
- obesity
- alcohol: current alcohol consumption
- age: age at onset
- chd: response, coronary heart disease.

```{r sahd,echo=FALSE, fig.cap="South African Heart Disease Data"}
# data source: https://web.stanford.edu/~hastie/ElemStatLearn/
dat <-  read.table("http://www-stat.stanford.edu/~tibs/ElemStatLearn/datasets/SAheart.data",sep=",",head=T,row.names=1)
vars <- c("chd","sbp","tobacco","ldl","famhist","obesity","alcohol","age")
dat <- dat[,vars]

pairs(data.matrix(dat[,-1]),col=ifelse(dat$chd==1,"red","blue"))
```

We fit a logitstic-regression model. 

```{r}
fit.logistic <- glm(chd~sbp+tobacco+ldl+famhist+obesity+alcohol+age,data=dat,family="binomial")
kable(broom::tidy(fit.logistic),digits=3)
```


There are some surprises in this table of coefficients, which must be interpreted with caution. Systolic blood pressure (sbp) is not significant! Nor
is obesity, and its sign is negative. This confusion is a result of the correlation between the set of predictors. On their own, both sbp and obesity
are significant, and with positive sign. However, in the presence of many other correlated variables, they are no longer needed (and can even get a
negative sign).

We proceed with backward selection.

```{r, echo=FALSE}
fit.bw <- stepAIC(fit.logistic,direction = "backward",trace=FALSE)
```

```{r}
kable(as.data.frame(fit.bw$anova),digits=3)
```

And the final regression coefficients are

```{r}
kable(broom::tidy(fit.bw),digits=3)
```

How does one interpret a coefficient of $0.081$ (Std. Error = $0.026$) for
tobacco, for example? Tobacco is measured in total lifetime usage in kilograms, with a median of $1.0$kg for the controls and $4.1$kg for the cases. Thus
an increase of 1kg in lifetime tobacco usage accounts for an increase in the
odds of coronary heart disease of $exp(0.081) = 1.084$ or $8.4$%. Incorporating the standard error we get an approximate $95$% confidence interval of
$exp(0.081 ± 2 × 0.026) = (1.03, 1.14)$.

Finally we run the Lasso approach

```{r}
x <- scale(data.matrix(dat[,-1]))
y <- dat$chd
fit.lasso <- glmnet(x=x,y=y,family="binomial")
plot(fit.lasso,xvar = "lambda",label=TRUE)
```

The first coefficient which is shrunk to zero is alcohol followed by obesity and sbp. 

We now turn to a truly high-dimensional example.

### Leukemia Dataset

<!--  Example from https://cran.r-project.org/web/packages/varbvs/vignettes/leukemia.html -->

The data consists of expression levels recorded for $3,571$ genes in $72$ patients with leukemia (Golub et al, 1999). The binary outcome encodes the disease subtype: acute lymphobastic leukemia (ALL) or acute myeloid leukemia (AML). The data are represented as a 72 x 3,571 matrix $x$ of gene expression values, and a vector $y$ of 72 binary disease outcomes. We first create train and test data.

```{r, echo=FALSE}
# set seed
set.seed(15)

# get leukemia data
library(varbvs) 
data(leukemia)
x <- leukemia$x
y <- leukemia$y

# test/train 
ind_train <- sample(1:length(y),size=length(y)/2)
xtrain <- x[ind_train,]
ytrain <- y[ind_train]
xtest <- x[-ind_train,]
ytest <- y[-ind_train]
```

We now run the elastic net

```{r}
# set settings for glmnet 
nfolds <- 10                  # number of cross-validation folds.
alpha  <- 0.95                # elastic net mixing parameter.

# run glmnet
fit.glmnet <-glmnet(xtrain,ytrain,family = "binomial",alpha=alpha)

# run cv.glmnet
cv.glmnet <- cv.glmnet(xtrain,ytrain,family = "binomial",type.measure = "class",
                       alpha = alpha,nfolds = nfolds)
```

The following Figure shows the Elastic net solution for the grid of lambda values.

```{r, include=FALSE}
plot(fit.glmnet,xvar="lambda",label=TRUE)
```

We run 10-fold cross-validation and take $\lambda_{\rm opt}$ as the largest $\lambda$ within 1 standard error of the minimum classification error.

```{r}
plot(cv.glmnet)
```

```{r}
(lambda.opt <- cv.glmnet$lambda.1se)
```

We extract the optimal coefficients and plot them as a barplot.

```{r}
beta.glmnet <- coef(fit.glmnet, s = cv.glmnet$lambda.1se)
barplot(as.numeric(beta.glmnet)[-1],xlab="gene",ylab="beta glment")
```

Finally we predict the disease outcome of the test samples using the fitted model and compare against the observed outcomes of the test samples.
```{r}
pred <- c(predict(fit.glmnet,xtest,s = lambda.opt,type = "class"))
print(table(true = factor(ytest),pred = factor(pred)))
```
The missclassification error on the test data is

```{r}
round(mean(pred!=ytest),3)
```

## Survival analysis

Another important application is survival analysis which deals with so-called time-to-event endpoints. We will give a very brief introduction into survival analysis and show how the cox proportional hazards model can be used to address high-dimensional problems. 

For subject $i$ we denote the event time with $T_i$. Unfortunately, typically we do not for each subject observe the event time as a subject may be censored due to:

* Loss to follow-up
* Withdrawal from study
* No event by end of fixed study period.

Therefore in practise we observe the survival time $\tilde{T}_i$ (which equals the event time or the censoring time whichever occurs earlier) and the event indicator $\delta_i$ ($\delta_i=1$ in case of event $=0$ in case of censoring).

A fundamental quantity in survival analysis is the survival probability

\[S(t)=P(T>t)=1-F(t).\]

The Kaplan-Meier method is the most common way to estimate survival times and probabilities.

The Cox proportional hazards model is commonly used for the study of the relationship between predictor variables and survival time. The Cox model assumes a semi-parametric form for the hazard, or the instantaneous rate at which events occur

\[h(t|X)=h_0(t)exp(X^T\beta).\]

$h_0(t)$ is the baseline hazard and $\beta$ are the regression coefficients. In the classic setting $n>p$, inference is made via the partial likelihood.

If you are interested to learn more about survival analysis I recommend you to have a look at this tutorial https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html.


### Regularized Cox Regression

In a high-dimension setting classical cox-regression falls short. Similar as in the previous sections we can use subset selection or use regularization. The latter is implemented in the function glmnet which penalizes the negative log of the partial likelihood with an elastic net penalty.


For illustration we consider the Lymphoma dataset which consists of gene expression data for $p=7399$ genes measured on $n=240$ patients, as well as censored survival times, for these patients. We start by reading the data and plotting the estimated survival curve.

```{r, include=FALSE}
# data source: http://web.stanford.edu/~hastie/StatLearnSparsity/data.html
# load packages for survival analysis
library(survival)
library(survminer)
set.seed(1)
```

```{r}
# read data
x <- read.table("data/lymphx.txt")%>%# gene expression (covariates)
  as.matrix
topvar.genes <- order(apply(x,2,var),decreasing=TRUE)[1:500]
x <- scale(x[,topvar.genes])
y <- read.table("data/lymphtime.txt",header = TRUE)%>%# survival time and event variable
  as.matrix
dat <- data.frame(cbind(y,x))
```

```{r}
# kaplan meier estimate
fit.surv <- survfit(Surv(time, status) ~ 1, data = dat)
ggsurvplot(fit.surv)
```

For example we can estimate the probability of surviving beyond 10 years.

```{r}
summary(survfit(Surv(time, status) ~ 1, data = dat), times = 10)
```

In this example we are interested to find the genes which are most predictive for survival. In this example we only consider the top $p=500$ genes with highest variance. 


We use glmnet to perform L1-penalized cox regression.

```{r}
set.seed(1)
fit.coxnet <- glmnet(x, y, family = "cox",alpha=0.95)
plot(fit.coxnet,xvar="lambda")
```

To identify the optimal lambda we use cross-validation and choose Harrel's concordance measure as the loss function.

```{r}
cv.coxnet <- cv.glmnet(x,y,
                       family="cox",
                       type.measure="C",
                       alpha=0.95)
plot(cv.coxnet)
```
We obtain the optimal lambda which maximizes the C-index

```{r}
cv.coxnet$lambda.min
```
and show the magnitute of the non-zero coefficients which can be viewed as a measure of variable importance (note that we standardized the input covariates).

```{r, echo=FALSE}
dbeta <- data.frame(NULL)
dbeta <- data.frame(betahat=as.numeric(coef(fit.coxnet,s=cv.coxnet$lambda.min)))
dbeta$betaname <- colnames(x)
dbeta%>%
  dplyr::filter(betahat!=0)%>%
  ggplot(.,aes(x=reorder(betaname,abs(betahat)),y=abs(betahat)))+
  geom_bar(stat="identity",width=0.25)+
  xlab("gene")+ylab("abs(betahat)")+
  theme(axis.text.x = element_text(angle = 45, hjust =1,size=5))
```

### Brier score and predictions in survival analysis

Evaluating the predictive performance of surivial models is an important topic. We consider here the prediction error defined as the time-dependent Brier score:

\[\rm{BS}(t,\hat{S})={\bf E}(Y_{\rm{new}}(t)-\hat{S}(t|X_{\rm new})), \]

where $Y_{\rm{new}}(t)={\bf 1}(T_{\rm new}\geq t)$ is the true status of a new test subject and
$\hat{S}(t|X_{\rm new})$ is the predicted survival probability. Calculation of the Brier score is not trivial as the we typically do not observe the event time $T$ for each subject (censoring). The R package pec estimates the Brier score using a technique called Inverse Probability of Censoring Weighting (IPCW). 

Let's illustrate the approach. We split the data set into train and test data.

```{r}
train_ind <- sample(1:nrow(x),size=nrow(x)/2)
xtrain <- x[train_ind,]
ytrain <- y[train_ind,]
xtest <- x[-train_ind,]
ytest <- y[-train_ind,]
dtrain <- data.frame(cbind(ytrain,xtrain))
dtest <- data.frame(cbind(ytest,xtest))
```

For illustration we use forward selection on the training data to obtain a prediction model.

```{r}
fit.lo <- coxph(Surv(time,status)~1,data=dtrain,
                x=TRUE,y=TRUE)
up <- as.formula(paste("~", 
                       paste(colnames(xtrain), 
                             collapse="+")))
fit.fw <- stepAIC(fit.lo,
                  scope=list(lower=fit.lo,
                             upper=up),
                  direction="forward",
                  steps=5,
                  trace=FALSE)

```

The following table summarizes the variables added in each step.forward selection approach.
```{r}
kable(as.data.frame(fit.fw$anova),digits=3)
```

The forest plot illustrates the estimated hazard ratios and confidence intervals.
```{r}
survminer::ggforest(fit.fw)
```

Finally we use the pec package to calculate the Brier score on the training and test data.
```{r, include=FALSE}
library(pec)
fit.pec.train <- pec::pec(
  object=list("cox.fw"=fit.fw), 
  data = dtrain, 
  formula = Surv(time, status) ~ 1, 
  splitMethod = "none")


fit.pec.test <- pec::pec(
  object=list("cox.fw"=fit.fw), 
  data = dtest, 
  formula = Surv(time, status) ~ 1, 
  splitMethod = "none")

```

Brier scores evaluted on training data.
```{r}
plot(fit.pec.train)
```
Brier score evaluated on test data
```{r}
plot(fit.pec.test)
```

We note that the selected model performs not better than the reference model (no covariates).

The pec package can also be used to benchmark different prediction models. We illustrate this based on random survival forest and backward selection. In this example we will not split the data into train and test. Instead we use cross-validation to compare the two prediction approaches.

We focus on only the variables with non zero coefficients in the glmnet fit.

```{r}
shat <- dbeta%>%
  dplyr::filter(betahat!=0)%>%
  pull(betaname)
fm <- as.formula(paste0("Surv(time, status) ~ ",paste(shat,collapse = "+")))
```


We run the random survival forest.
```{r}
fit.cforest <- pec::pecCforest(fm, data =dat, 
                              control = party::cforest_classical(ntree = 1000))

```

Similar as for classical random forest we can obtain the variable importance.

```{r, eval=FALSE}
vp <- party::varimp(fit.cforest$forest)
dvp <- data.frame(VARNAME=names(vp),VARIMP=vp)
vplot <- dvp%>%
  arrange(desc(VARIMP))%>%
  ggplot()+
  geom_point(aes(x = reorder(VARNAME, VARIMP), y = VARIMP), size = 3, shape = 18)+
  labs(title = "Variable Importance Random Survival Forest")+
  xlab("Variable")+
  ylab("Variable importance")+
  coord_flip()
print(vplot)
```

Next we run backward selection.

```{r}
fit.coxbw <- pec::selectCox(fm, data = dat, rule = "aic")
```

And we compare the prediction performance of the two approaches using the cross-validated Brier score. 

```{r message=FALSE}
pec.cv <- pec::pec(
  object=list("cox.bw"=fit.coxbw,"cforest"=fit.cforest), 
  data = dat, 
  formula = Surv(time, status) ~ 1, 
  splitMethod = "cv10")
plot(pec.cv)
```


