# Classification and Survival Analysis

```{r include=FALSE}
library(knitr)
library(caret)
library(mlbench)
library(glmnet)
library(tidyverse)
library(e1071)
library(MASS)
```

We so far focused on the linear regression model. We will now extend to the setting where the response is binary or time-to-event. The former is also referred to as classification.

## Classification

We focus on the binary situation where response $G$ takes values $0$ and $1$, and, the aim is to do prediction based based on covariates $X=(X_1,\ldots,X_p)$. We model the probability of success $p(x;\beta)=P(G=1|X=x;\beta)$ using the logistic regression approach with logit link function

\[\rm{logit}(x;\beta)=log \Big(\frac{p(x;\beta)}{1-p(x;\beta)}\Big)=X^T\beta.\]

In logistic regression we estimate the regression parameter $\beta$ by maximizing the log-likelihood

\begin{align*}
\ell(\beta|\bf G,\bf X)&=\sum_{i=1}^{n}(1-g_i)\log(1-p(x_i;\beta))+g_i\log p(x_i;\beta)\\
&=\sum_{i=1}^{n}g_i x_i^T\beta - \log(1+\exp(x_i^T\beta)).
\end{align*}

Given an estimate $\hat \beta$ (from training data), prediction based on new input data $X_{\rm new}$ can be obtained via the predicted probablity of success $p(X_{\rm new};\hat \beta)$, e.g. the class labels corresponding to the maximum probability

\begin{align*}
  \hat{G}(X_{\rm new})=\left\{
    \begin{array}{ll}
      1, & \mbox{if $p(X_{\rm new};\hat\beta)>0.5$}.\\
      0, & \mbox{otherwise}.
    \end{array}
  \right.
\end{align*}

Similar as in linear regression, in the high-dimensional context where $n$ is small compared to $p$, the maximum likelihood estimator does often lead to overfitting and a poor generalisation error. We will in the following generalize the previously introduced idea of regularization to classification. In the context of linear regression we penalized the residual sum of squares (RSS). We extend this and penalize the negative log-likelihood with an elastic net penalty

\begin{align*}
\hat{\beta}_{\alpha,\lambda}&=\rm{argmin}-\frac{1}{n}\ell(\beta|\bf G,\bf X)+\lambda\big((1-\alpha)\|\beta\|_2^2/2+\alpha\|\beta\|_1\big).
\end{align*}

There are different measures to judge the quality of the predictions. We focus on the missclassification error  which is simply the fraction of missclassified test samples. Another important measure used in the context of classification is the receiver operating characteristic (ROC). 

### South African Heart Disease Data

We start with a "low" dimensional example illustrating the logistic regression model and comparing 
the Lasso with traditional backward selection. The data in Figure 4.12 are a
subset of the Coronary Risk-Factor Study (CORIS) baseline survey, carried
out in three rural areas of the Western Cape, South Africa (Rousseauw et al, 1983, South African Medical
Journal). The aim of the study was to establish the intensity of ischemic
heart disease risk factors in that high-incidence region. The data represent
white males between 15 and 64, and the response variable is the presence or
absence of myocardial infarction (MI) at the time of the survey (the overall
prevalence of MI was 5.1% in this region). There are 160 cases in our data
set, and a sample of 302 controls. This example is taken from the book "Elements of Statistical Learning". The variables are:

- sbp: systolic blood pressure
- tobacco: cumulative tobacco (kg)
- ldl: low densiity lipoprotein cholesterol adiposity
- famhist: family history of heart disease (Present, Absent)
- obesity
- alcohol: current alcohol consumption
- age: age at onset
- chd: response, coronary heart disease.

```{r,echo=FALSE}
# data source: https://web.stanford.edu/~hastie/ElemStatLearn/
dat <-  read.table("http://www-stat.stanford.edu/~tibs/ElemStatLearn/datasets/SAheart.data",sep=",",head=T,row.names=1)
vars <- c("chd","sbp","tobacco","ldl","famhist","obesity","alcohol","age")
dat <- dat[,vars]

pairs(data.matrix(dat[,-1]),col=ifelse(dat$chd==1,"red","blue"))
```

We fit a logitstic-regression model. 
There are some surprises in this table of coefficients, which must be in-
terpreted with caution. Systolic blood pressure (sbp) is not significant! Nor
is obesity, and its sign is negative. This confusion is a result of the corre-
lation between the set of predictors. On their own, both sbp and obesity
are significant, and with positive sign. However, in the presence of many other correlated variables, they are no longer needed (and can even get a
negative sign).

```{r}
fit.logistic <- glm(chd~sbp+tobacco+ldl+famhist+obesity+alcohol+age,data=dat,family="binomial")
kable(broom::tidy(fit.logistic),digits=3)
```

We proceed with backward selection.

```{r, echo=FALSE}
fit.bw <- stepAIC(fit.logistic,direction = "backward",trace=FALSE)
```

```{r}
kable(as.data.frame(fit.bw$anova),digits=3)
```

And the final regression coefficients are

```{r}
kable(broom::tidy(fit.bw),digits=3)
```

How does one interpret a coefficient of $0.081$ (Std. Error = $0.026$) for
tobacco, for example? Tobacco is measured in total lifetime usage in kilo-
grams, with a median of $1.0$kg for the controls and $4.1$kg for the cases. Thus
an increase of 1kg in lifetime tobacco usage accounts for an increase in the
odds of coronary heart disease of $exp(0.081) = 1.084$ or $8.4$%. Incorporat-
ing the standard error we get an approximate $95$% confidence interval of
$exp(0.081 ± 2 × 0.026) = (1.03, 1.14)$.

Finally we run the Lasso approach

```{r,echo=FALSE}
x <- scale(data.matrix(dat[,-1]))
y <- dat$chd
fit.lasso <- glmnet(x=x,y=y,family="binomial")
plot(fit.lasso,xvar = "lambda",label=TRUE)
```

The first coefficient which is shrunk to zero is alcohol followed by obesity and sbp. 

We now turn to a truly high-dimensional example.

### Leukemia Dataset

<!--  Example from https://cran.r-project.org/web/packages/varbvs/vignettes/leukemia.html -->

The data consists of expression levels recorded for $3,571$ genes in $72$ patients with leukemia (Golub et al, 1999). The binary outcome encodes the disease subtype: acute lymphobastic leukemia (ALL) or acute myeloid leukemia (AML). The data are represented as a 72 x 3,571 matrix x of gene expression values, and a vector y of 72 binary disease outcomes.

```{r, echo=FALSE}
# set seed
set.seed(15)

# get leukemia data
library(varbvs) 
data(leukemia)
x <- leukemia$x
y <- leukemia$y

# test/train 
ind_train <- sample(1:length(y),size=length(y)/2)
xtrain <- x[ind_train,]
ytrain <- y[ind_train]
xtest <- x[-ind_train,]
ytest <- y[-ind_train]
```

We now run the elastic net

```{r}
# set settings for glmnet 
nfolds <- 10                  # number of cross-validation folds.
alpha  <- 0.95                  # elastic net mixing parameter.

# run glmnet
fit.glmnet <-glmnet(xtrain,ytrain,family = "binomial",alpha=alpha)

# run cv.glmnet
cv.glmnet <- cv.glmnet(xtrain,ytrain,family = "binomial",type.measure = "class",
                       alpha = alpha,nfolds = nfolds)
```


The following Figure shows the Elastic net solution for the grid of lambda values.

```{r}
plot(fit.glmnet,xvar="lambda",label=TRUE)
```

We run 10-fold cross-validation and take $\lambda_{\rm opt}$ as the largets $\lambda$ within 1 standard error of the minimum classification error.

```{r}
plot(cv.glmnet)
lambda.opt <- cv.glmnet$lambda.1se
```

The optimal coefficients can be optained as

```{r, echo=FALSE}
beta.glmnet <- coef(fit.glmnet, s = cv.glmnet$lambda.1se)
barplot(as.numeric(beta.glmnet)[-1],xlab="gene",ylab="beta glment")
```

Finally we predict the disease outcome using the fitted model and compare against the observed values
```{r}
pred <- c(predict(fit.glmnet,xtest,s = lambda.opt,type = "class"))
print(table(true = factor(ytest),pred = factor(pred)))
```
and the missclassification error is

```{r}
round(mean(pred!=ytest),3)
```


### Sonar Dataset

To illustrate the approaches we use the Sonar dataset from the mlbench package. The goal is to train a model to discriminate between rock "R" and metal "M" based on sonar signals bounced off a metal cylinder and those bounced off a roughly cylindrical rock. The $p=60$ covariates represent energies within particular frequency bands.

```{r}
data(Sonar)

set.seed(107)
inTrain <- createDataPartition(
  y = Sonar$Class,
  ## the outcome data are needed
  p = .5,
  ## The percentage of data in the
  ## training set
  list = FALSE
)
dtrain <- Sonar[ inTrain,]
xtrain <- as.matrix(dtrain[,-61])
ytrain <- dtrain$Class
dtest  <- Sonar[-inTrain,]
xtest <- as.matrix(dtest[,-61])
ytest <- dtest$Class
```


We first run logistic regression once with only an intercept and once with all the covariates. We
note that glm reports convergence problems when using all variables. This indicates that the model could be to complex to fit.

In order to compare the performance of the different methods we setup trainControl. We use 10 fold cross-validation

```{r}
tc <- trainControl(method = "cv", number = 10) # 10-fold CV, performance=accuracy

# tc <- trainControl(method = "cv", number = 10, # 10-fold CV, performance=ROC
#                    summaryFunction = twoClassSummary,
#                    classProbs = TRUE)
```


```{r}
fit.0 <-  glm(Class~1,family=binomial,data=dtrain)
fit.full <-  glm(Class~.,family=binomial,data=dtrain)
fit.full.caret <- train(Class ~., 
                        data = dtrain, 
                        method = "glm", 
                        family = "binomial",
                        trControl = tc)
```

Next we use forward selection.

```{r}
upper.model <- paste("~", paste(colnames(dtrain), collapse=" + ")) 
fit.fw <- stepAIC(fit.0,
                  scope=upper.model,
                  direction = "forward", 
                  steps=10,
                  trace = FALSE)
summary(fit.fw)
broom::tidy(fit.fw)
```


```{r}
fit.fw.caret <- train(x=xtrain,
                      y=ytrain, 
                      method = "glmStepAIC", 
                      family = "binomial",
                      direction ="forward",
                      steps=10,
                      trControl = tc,
                      trace = FALSE
)
# accuracy
fit.fw.caret$results
# summary of the model
summary(fit.fw.caret$finalModel)
```

Next, we use L1-penalized regression.


```{r}
fit.l1 <-glmnet(x=xtrain,y=ytrain,alpha=1,family="binomial") 
plot(fit.l1,xvar="lambda",label=TRUE)
cvfit <- cv.glmnet(x=xtrain, y=ytrain, family = "binomial", type.measure = "class",alpha=1)
cvfit$lambda.min
plot(cvfit)
```

```{r}
lambda.grid <- fit.l1$lambda
fit.l1.caret<-train(x=xtrain,
                 y=ytrain,
                 method = "glmnet",
                 family="binomial",
                 preProc = c("center","scale"),
                 tuneGrid = expand.grid(alpha = 1,lambda=lambda.grid),
                 trControl = tc
) 

#
ggplot(fit.l1.caret)
# Best lambda
fit.l1.caret$bestTune$lambda
# Model coefficients
coef(fit.l1.caret$finalModel,fit.l1.caret$bestTune$lambda)
# Make predictions
confusionMatrix(data =fit.l1.caret %>% predict(xtest), ytest)
confusionMatrix(data =fit.fw.caret %>% predict(xtest), ytest)
confusionMatrix(data =fit.full.caret %>% predict(xtest), ytest)
```

Finally we compare all models
```{r}
models <- list(full=fit.full.caret,
               fw=fit.fw.caret,
               l1=fit.l1.caret)
summary(resamples(models),metric = "Accuracy")
```

```{r}
coef(fit.l1,s=cvfit$lambda.min)
coef(fit.fw)
```


## Survival analysis

Another important endpoint in many areas is time-to-event. We will give a very brief introduction into survival analysis and show the cox proportional hazards model can be used to address high-dimensional problems. For subject $i$ we denote the event time with $T_i$. Unfortunately, typically we do for each subject observe the event time as a subject may be censored due to:

* Loss to follow-up
* Withdrawal from study
* No event by end of fixed study period.

Therefore in practise we observe the survival time $Y_i$ (which e) and the event indicator $\delta_i$ ($\delta_i=1$ in case of event $=0$ in case of censoring).

A fundamental quantity in survival analysis is the survival probability

\[S(t)=P(T>t)=1-F(t).\]

The Kaplan-Meier method is the most common way to estimate survival times and probabilities.
The survival probability can be estimated using the Kaplan-Meier method.


The Cox proportional hazards model is commonly used for the study of the relationship beteween predictor variables and survival time. The Cox model assumes a semi-parametric form for the hazard, or the instantaneous rate at which events occur

\[h(t|X)=h_0(t)exp(X^T\beta).\]

$\lambda_0(t)$ is the baseline hazard and $\beta$ are the regression coefficients. In the classic setting $n>p$, inference is made via the partial likelihood. In the high-dimensional context we can penalize the negative log of the partial likelihood.

If you are interested to learn more about survival analysis I recommend you to have a look at this tutorial https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html.


### Example Dataset
Gene expression data for 7399 genes measured on 240 patients, as well as censored survival times, for these patients. We start by reading the data and plotting the estimated surival curve.

```{r}
# data source: http://web.stanford.edu/~hastie/StatLearnSparsity/data.html
# load packages for survival analysis
library(survival)
library(survminer)

# read data
x <- read.table("data/lymphx.txt")%>%# gene expression (covariates)
  as.matrix
y <- read.table("data/lymphtime.txt",header = TRUE)%>%# survival time and event variable
  as.matrix
dat <- data.frame(cbind(y,x))

# kaplan meier estimate
fit.surv <- survfit(Surv(time, status) ~ 1, data = dat)
ggsurvplot(fit.surv)
```

For example the estimated probability of surviving beyond 10 years is obtained as

```{r}
summary(survfit(Surv(time, status) ~ 1, data = dat), times = 10)
```

We now perform L1-penalized cox regression using glmnet.

```{r}
set.seed(1)
fit.coxnet <- glmnet(x, y, family = "cox",alpha=1)
plot(fit.coxnet,xvar="lambda")

cv.coxnet <- cv.glmnet(x,y,family="cox")
plot(cv.coxnet)
#plot(cv.coxnet, xlim = c(log(0.05), log(0.5)), ylim = c(11.3,12.4))

cv.coxnet$lambda.min

betahat <- coef(fit.coxnet,s=cv.coxnet$lambda.min)
Shat <- rownames(betahat)[which(betahat != 0)]

fm <- as.formula(paste0("Surv(time, status) ~ 0+",paste(Shat,collapse = "+")))
fit.shat <- coxph(fm,data=dat,x=TRUE,y=TRUE)
```



```{r, eval=FALSE}
library(party)
library(randomForestSRC)
# Condicional Inferences Trees
fm <- Surv(time, status) ~ .
fitcforest <- pec::pecCforest(fm, data =dat[,1:20], 
                              control = party::cforest_classical(ntree = 1000))

# Survival Random Forests
fitrsf <- randomForestSRC::rfsrc(fm, data = dat[,1:20], forest = TRUE, 
                                 ntree = 1000, splitrule = "logrank", 
                                 na.action = "na.impute")

# Variable Importance
varimp <- randomForestSRC::vimp(fitrsf, importance = "permute")$importance
data.frame(varimp)%>%
  tibble::rownames_to_column("NAME")%>%
  ggplot()+
  geom_point(aes(x = reorder(NAME, varimp), y = varimp), size = 3, shape = 18)+
  labs(title = "Variable Importance Random Survival Forest")+
  xlab("Variable")+
  ylab("Variable importance")+
  coord_flip()
```

```{r, eval=FALSE}
library(pec)
models <- list("cox.shat"=fit.shat)
fitpec <- pec::pec(
  object=models, 
  data = dat, 
  formula = Surv(time, status) ~ 1, 
  splitMethod = "cv10", B = 1, keep.index = TRUE, keep.matrix = TRUE)
plot(fitpec)
```



