# Extension to Survival Analysis

```{r include=FALSE}
library(knitr)
library(glmnet)
library(e1071)
library(MASS)
library(party)
library(pec)
library(tidyverse)
```

We turn our attention to survival analysis which deals with so-called time-to-event endpoints. We start with a very brief introduction to survival analysis and introduce the cox proportional hazards model. Then we turn to the high-dimensional setting and discuss elastic net regularization for cox regression. Finally, we introduce the time-dependent Brier score, a popular prediction performance measure in the survival setting. 

## Survival analysis and proportional hazard model

For subject $i$ we denote the event time with $T_i$. Typically we do not for each subject observe the event time as a subject may be censored due to:

* Loss to follow-up
* Withdrawal from study
* No event by end of fixed study period.

Therefore in practise we observe the survival time $\widetilde{T}_i$ (which equals the event time or the censoring time whichever occurs earlier) and the event indicator $\delta_i$ ($\delta_i=1$ in case of event $=0$ in case of censoring).

A fundamental quantity in survival analysis is the survival probability

\[S(t)=P(T>t)=1-F(t).\]

The Kaplan-Meier method is the most common way to estimate survival times and probabilities in the context of censoring. 


Another important quantitiy is the hazard function $h(t)$, or the instantaneous rate at which events occur, which is defined as

\[h(t)=\lim_{dt\rightarrow 0}\frac{P(t\leq T < t+dt|T\geq t)}{dt}=-S'(t)/S(t).\]

The Cox proportional hazards model is commonly used for the study of the relationship between predictor variables and survival time. The Cox model assumes a semi-parametric form for the hazard

\[h(t|X)=h_0(t)\exp(X^T\beta).\]

$h_0(t)$ is the baseline hazard and $\beta$ are the regression coefficients. In the classic cox regression estimation is done by optimizing the log partial likelihood $\ell(\beta|{\bf \widetilde{T} ,\Delta,X})$.

To learn more about survival analysis I recommend to have a look at this short tutorial https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html.


## Regularized Cox Regression

In a high-dimension setting classical cox regression falls short. Similar as in the previous sections we can use subset selection or regularization. The R package `glmnet` implements elastic-net penalized cox regression.

For illustration we use the Lymphoma data set which consists of gene expression data for $p=7399$ genes measured on $n=240$ patients, as well as censored survival times, for these patients. We start by reading the data.

```{r, include=FALSE}
# data source: http://web.stanford.edu/~hastie/StatLearnSparsity/data.html
# load packages for survival analysis
library(survival)
library(survminer)
set.seed(1)
```

```{r}
# read gene expression matrix
x <- read.table("data/lymphx.txt")%>%
  as.matrix
topvar.genes <- order(apply(x,2,var),decreasing=TRUE)[1:500]
x <- scale(x[,topvar.genes])

# read survival time and event variable
y <- read.table("data/lymphtime.txt",header = TRUE)%>%
  as.matrix
```
The first column of `y` is the survival time and the second column the event indicator.

```{r}
head(y)
```

Next we plot the estimated survival curve.

```{r}
dat <- data.frame(cbind(y,x))
fit.surv <- survfit(Surv(time, status) ~ 1, data = dat)
ggsurvplot(fit.surv)
```

We can use the `summary` function to obtain more specific information on estimated survival quantities, e.g. the probability of surviving beyond 10 years is obtain as follows. 

```{r, eval=FALSE}
summary(survfit(Surv(time, status) ~ 1, data = dat), times = 10)
```

We are interested to find the genes which are most predictive for survival. We assume proportionality of hazards and use `glmnet` to perform L1-penalized cox regression. (Note that we only consider $p=500$ genes with largest variance.)

```{r}
set.seed(1)
fit.coxnet <- glmnet(x, y, family = "cox",alpha=0.95)
plot(fit.coxnet,xvar="lambda")
```

To identify the optimal lambda we use cross-validation and take Harrel's concordance index as a goodness of fit measure. 

```{r}
cv.coxnet <- cv.glmnet(x,y,
                       family="cox",
                       type.measure="C",
                       alpha=0.95)
plot(cv.coxnet)
```

The C-index ranges from 0.5 to 1. A value of 0.5 means that the model is no better than predicting an outcome than random chance. The tuning parameter which maximizes the C-index is $\lambda_{\rm{opt}}=$ `r round(cv.coxnet$lambda.min,2)`. The next graphic shows the magnitude of the non-zero coefficients (note that we standardized the input covariates).

```{r, echo=FALSE}
dbeta <- data.frame(NULL)
dbeta <- data.frame(betahat=as.numeric(coef(fit.coxnet,s=cv.coxnet$lambda.min)))
dbeta$betaname <- colnames(x)
dbeta%>%
  dplyr::filter(betahat!=0)%>%
  ggplot(.,aes(x=reorder(betaname,abs(betahat)),y=abs(betahat)))+
  geom_bar(stat="identity",width=0.25)+
  xlab("gene")+ylab("abs(betahat)")+
  theme(axis.text.x = element_text(angle = 45, hjust =1,size=7),
        text = element_text(size = 15))
```

## Brier score and predictions in survival analysis

Evaluating the predictive performance in the linear regression and classification context was straightforward. For time-to-event data this is more involved. A popular score to measure the prediction accuracy is the time-dependent Brier score

\[{\rm BS}(t,\hat{S})={\bf E}(Y_{\rm{new}}(t)-\hat{S}(t|X_{\rm new})) \]

where $Y_{\rm{new}}(t)={\bf 1}(T_{\rm new}\geq t)$ is the true status of a new test subject and
$\hat{S}(t|X_{\rm new})$ is the predicted survival probability. Calculation of the Brier score is complicated by the fact that we do not always observe the event time $T$ due to censoring. The R package `pec` estimates the Brier score using a technique called *Inverse Probability of Censoring Weighting (IPCW)*. 

We split the data set into training and test data.

```{r}
train_ind <- sample(1:nrow(x),size=nrow(x)/2)
xtrain <- x[train_ind,]
ytrain <- y[train_ind,]
xtest <- x[-train_ind,]
ytest <- y[-train_ind,]
dtrain <- data.frame(cbind(ytrain,xtrain))
dtest <- data.frame(cbind(ytest,xtest))
```

For illustration we use forward selection on the training data to obtain a prediction model.

```{r}
fit.lo <- coxph(Surv(time,status)~1,data=dtrain,
              x=TRUE,y=TRUE)
up <- as.formula(paste("~", 
                       paste(colnames(xtrain), 
                             collapse="+")))
fit.fw <- stepAIC(fit.lo,
                  scope=list(lower=fit.lo,
                             upper=up),
                  direction="forward",
                  steps=5,
                  trace=FALSE)

```

The following table summarizes the variables added in each step of the forward selection approach.
```{r}
kable(as.data.frame(fit.fw$anova),digits=3)
```

The next figure shows the estimated hazard ratios and confidence intervals. This type of plot is often referred to as forest plot.

```{r, warning=FALSE}
survminer::ggforest(fit.fw)
```

Finally we use the `pec` package to calculate the Brier score on the training and test data.

```{r message=FALSE}
library(pec)
fit.pec.train <- pec::pec(
  object=list("cox.fw"=fit.fw), 
  data = dtrain, 
  formula = Surv(time, status) ~ 1, 
  splitMethod = "none")


fit.pec.test <- pec::pec(
  object=list("cox.fw"=fit.fw), 
  data = dtest, 
  formula = Surv(time, status) ~ 1, 
  splitMethod = "none")

```
The following figure shows the Brier scores evaluated on training and test data.
```{r}
par(mfrow=c(1,2))
plot(fit.pec.train,main="training data")
plot(fit.pec.test,main="test data")
```

The plot on the right shows the Brier score on the test data and indicates that the selected model performs no better than the reference model (no covariates).

The `pec` package can also be used to benchmark different prediction models. We illustrate this based on random forest and backward selection. In this illustration we do not split the data into training and test. Instead we use cross-validation to compare the two prediction approaches. We focus on only the covariates which have non-zero coefficient in the previous glmnet fit.

```{r}
shat <- dbeta%>%
  dplyr::filter(betahat!=0)%>%
  pull(betaname)
fm <- as.formula(paste0("Surv(time, status) ~ ",paste(shat,collapse = "+")))
```

We start by fitting random forest using `cforest` from the `party` package.

```{r}
fit.cforest <- pec::pecCforest(fm, data =dat, 
                              control = party::cforest_classical(ntree = 100))
```

We can obtain a measure of variable importance using the function `varimp`.

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=8}
vp <- party::varimp(fit.cforest$forest)
dvp <- data.frame(VARNAME=names(vp),VARIMP=vp)
vplot <- dvp%>%
  arrange(desc(VARIMP))%>%
  ggplot()+
  geom_point(aes(x = reorder(VARNAME, VARIMP), y = VARIMP), size = 3, shape = 18)+
  labs(title = "Variable Importance Random Survival Forest")+
  xlab("Variable")+
  ylab("Variable importance")+
  coord_flip()
print(vplot)
```

Next we run Backward selection.

```{r}
fit.coxbw <- pec::selectCox(fm, data = dat, rule = "aic")
```

Finally we compare the two approaches using the cross-validated Brier score. 

```{r message=FALSE, warning=FALSE}
pec.cv <- pec::pec(
  object=list("cox.bw"=fit.coxbw,"cforest"=fit.cforest), 
  data = dat, 
  formula = Surv(time, status) ~ 1, 
  splitMethod = "cv10")
plot(pec.cv)
```

We conclude that random forest exhibits a slightly better prediction performance.
