---
title: "Classification and Survival Analysis"
author: "Nicolas Städler"
date: "3/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results="hide",warnings=FALSE)
```


```{r, echo=FALSE}
library(caret)
library(mlbench)
library(glmnet)
library(tidyverse)
library(e1071)
library(MASS)
```

We so far focused on linear regression. We will now extend to situation where response is binary (classification) and where response is time-to-event. 

## Classification

We focus on the binary situation where response $G$ takes values $0$ and $1$, and, the aim is to do prediction based based on covariates $X=(X_1,\ldots,X_p)$. We model the probability of success $p(x)=P(G=1|X=x;\beta)$ using the logistic regression approach with logit link function

\[\rm{logit}(x;\beta)=log \Big(\frac{p(x;\beta)}{1-p(x;\beta)}\Big)=X^T\beta.\]

In logistic regression we estimate the regression parameter $\beta$ by maximizing the log-likelihood

\begin{align*}
l(\beta|\bf G,\bf X)&=\sum_{i=1}^{n}(1-g_i)\log(1-p(x_i;\beta))+g_i\log p(x_i;\beta)\\
&=\sum_{i=1}^{n}g_i x_i^T\beta - \log(1+\exp(x_i^T\beta)).
\end{align*}

Given an estimate $\hat \beta$ (from training data), prediction based on new input data $X_{\rm new}$ is obtained as

\begin{align*}
\hat{G}(X_{\rm new})&=1\quad \rm{if}\quad p(X_{\rm new};\hat\beta)>0.5. 
\end{align*}

The quality of the prediction is judged based on the missclassification error on test data.

Similar as in linear regression, in the high-dimensional context where $n$ is small compared to $p$ the maximum likelihood estimater (i.e. standard logistic regression) does often lead to overfitting and poor generalisation error. We will discuss in the following forward regression, L1-penalized likelihood and support vector machines as alternative methods.

### Example Dataset

To illustrate the approaches we use the Sonar dataset from the mlbench package. The goal is to train a model to discriminate between rock "R" and metal "M" based on sonar signals bounced off a metal cylinder and those bounced off a roughly cylindrical rock. The $p=60$ covariates represent energies within particular frequency bands.

```{r}
data(Sonar)

set.seed(107)
inTrain <- createDataPartition(
  y = Sonar$Class,
  ## the outcome data are needed
  p = .5,
  ## The percentage of data in the
  ## training set
  list = FALSE
)
dtrain <- Sonar[ inTrain,]
xtrain <- as.matrix(dtrain[,-61])
ytrain <- dtrain$Class
dtest  <- Sonar[-inTrain,]
xtest <- as.matrix(dtest[,-61])
ytest <- dtest$Class

```

### Logistic regression, Forward selection and L1 penalization

We first run logistic regression once with only an intercept and once with all the covariates. We
note that glm reports convergence problems when using all variables. This indicates that the model could be to complex to fit.

In order to compare the performance of the different methods we setup trainControl. We use 10 fold cross-validation

```{r}
tc <- trainControl(method = "cv", number = 10) # 10-fold CV, performance=accuracy

# tc <- trainControl(method = "cv", number = 10, # 10-fold CV, performance=ROC
#                    summaryFunction = twoClassSummary,
#                    classProbs = TRUE)
```


```{r}
fit.0 <-  glm(Class~1,family=binomial,data=dtrain)
fit.full <-  glm(Class~.,family=binomial,data=dtrain)
fit.full.caret <- train(Class ~., 
                        data = dtrain, 
                        method = "glm", 
                        family = "binomial",
                        trControl = tc)
```

Next we use forward selection.

```{r}
upper.model <- paste("~", paste(colnames(dtrain), collapse=" + ")) 
fit.fw <- stepAIC(fit.0,
                  scope=upper.model,
                  direction = "forward", 
                  steps=10,
                  trace = FALSE)
summary(fit.fw)
broom::tidy(fit.fw)
```


```{r}
fit.fw.caret <- train(x=xtrain,
                      y=ytrain, 
                      method = "glmStepAIC", 
                      family = "binomial",
                      direction ="forward",
                      steps=10,
                      trControl = tc,
                      trace = FALSE
)
# accuracy
fit.fw.caret$results
# summary of the model
summary(fit.fw.caret$finalModel)
```

Next, we use L1-penalized regression

\begin{align*}
\hat{\beta}^{\rm Lasso}_{\lambda}&=\rm{argmin}-l(\beta|\bf G,\bf X)+\lambda\|\beta\|.
\end{align*}


```{r}
fit.l1 <-glmnet(x=xtrain,y=ytrain,alpha=1,family="binomial") 
plot(fit.l1,xvar="lambda",label=TRUE)
cvfit <- cv.glmnet(x=xtrain, y=ytrain, family = "binomial", type.measure = "class",alpha=1)
cvfit$lambda.min
plot(cvfit)
```

```{r}
lambda.grid <- fit.l1$lambda
fit.l1.caret<-train(x=xtrain,
                 y=ytrain,
                 method = "glmnet",
                 family="binomial",
                 preProc = c("center","scale"),
                 tuneGrid = expand.grid(alpha = 1,lambda=lambda.grid),
                 trControl = tc
) 

#
ggplot(fit.l1.caret)
# Best lambda
fit.l1.caret$bestTune$lambda
# Model coefficients
coef(fit.l1.caret$finalModel,fit.l1.caret$bestTune$lambda)
# Make predictions
confusionMatrix(data =fit.l1.caret %>% predict(xtest), ytest)
confusionMatrix(data =fit.fw.caret %>% predict(xtest), ytest)
confusionMatrix(data =fit.full.caret %>% predict(xtest), ytest)
```

Finally we compare all models
```{r}
models <- list(full=fit.full.caret,
               fw=fit.fw.caret,
               l1=fit.l1.caret)
summary(resamples(models),metric = "Accuracy")
```

```{r}
coef(fit.l1,s=cvfit$lambda.min)
coef(fit.fw)
```


## Survival analysis

Another important endpoint in many areas is time-to-event. We will give a very brief introduction into survival analysis and show the cox proportional hazards model can be used to address high-dimensional problems. For subject $i$ we denote the event time with $T_i$. Unfortunately, typically we do for each subject observe the event time as a subject may be censored due to:

* Loss to follow-up
* Withdrawal from study
* No event by end of fixed study period.

Therefore in practise we observe the survival time $Y_i$ (which e) and the event indicator $\delta_i$ ($\delta_i=1$ in case of event $=0$ in case of censoring).

A fundamental quantity in survival analysis is the survival probability

\[S(t)=P(T>t)=1-F(t).\]

The Kaplan-Meier method is the most common way to estimate survival times and probabilities.
The survival probability can be estimated using the Kaplan-Meier method.


The Cox proportional hazards model is commonly used for the study of the relationship beteween predictor variables and survival time. The Cox model assumes a semi-parametric form for the hazard, or the instantaneous rate at which events occur

\[h(t|X)=h_0(t)exp(X^T\beta).\]

$\lambda_0(t)$ is the baseline hazard and $\beta$ are the regression coefficients. In the classic setting $n>p$, inference is made via the partial likelihood. In the high-dimensional context we can penalize the negative log of the partial likelihood.

If you are interested to learn more about survival analysis I recommend you to have a look at this tutorial https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html.


### Example Dataset
Gene expression data for 7399 genes measured on 240 patients, as well as censored survival times, for these patients. We start by reading the data and plotting the estimated surival curve.

```{r}
# load packages for survival analysis
library(survival)
library(survminer)

# read data
x <- read.table("data/lymphx.txt")%>%# gene expression (covariates)
  as.matrix
y <- read.table("data/lymphtime.txt",header = TRUE)%>%# survival time and event variable
  as.matrix
dat <- data.frame(cbind(y,x))

# kaplan meier estimate
fit.surv <- survfit(Surv(time, status) ~ 1, data = dat)
ggsurvplot(fit.surv)
```

For example the estimated probability of surviving beyond 10 years is obtained as

```{r}
summary(survfit(Surv(time, status) ~ 1, data = dat), times = 10)
```

We now perform L1-penalized cox regression using glmnet.

```{r}
fit.l1 <- glmnet(x, y, family = "cox",alpha=1)
plot(fit.l1,xvar="lambda")

set.seed(1)
cvfit <- cv.glmnet(x,y,family="cox",type.measure = "C")
plot(cvfit)

cvfit$lambda.min
cvfit$lambda.1se

betahat <- coef(fit.l1,s=cvfit$lambda.1se)
names(betahat) <- colnames(x)
Shat <- rownames(betahat)[which(betahat != 0)]

fm <- as.formula(paste0("Surv(time, status) ~ 0 +",paste(Shat,collapse = "+")))
fit.shat <- coxph(fm,data=dat,x=TRUE,y=TRUE)
```


library(pec)
# cox backward selection with adsl variables
fm <- Surv(AVAL, Delta) ~ 0 + A1 + A2 + A3+GROUP
fitcoxbw <- pec::selectCox(fm, data = dt_tte, rule = "aic")
print(survminer::ggforest(coxph(fitcoxbw$fit$sformula, data = dt_tte)))



```{r}
library(party)
# Condicional Inferences Trees
fm <- Surv(time, status) ~ .
fitcforest <- pec::pecCforest(fm, data =dat[,1:20], 
                              control = party::cforest_classical(ntree = 1000))

# Survival Random Forests
fitrsf <- randomForestSRC::rfsrc(fm, data = dt_tte, forest = TRUE, 
                                 ntree = 1000, splitrule = "logrank", 
                                 na.action = "na.impute")

# Variable Importance
varimp <- randomForestSRC::vimp(fitrsf, importance = "permute")$importance
data.frame(varimp)%>%
  tibble::rownames_to_column("NAME")%>%
  ggplot()+
  geom_point(aes(x = reorder(NAME, varimp), y = varimp), size = 3, shape = 18)+
  labs(title = "Variable Importance Random Survival Forest")+
  xlab("Variable")+
  ylab("Variable importance")+
  coord_flip()
```

```{r}
models <- list("cox.shat"=fit.shat)
fitpec <- pec::pec(
  object=models, 
  data = dat, 
  formula = Surv(time, status) ~ 1, 
  splitMethod = "cv10", B = 1, keep.index = TRUE, keep.matrix = TRUE)
plot(fitpec)
```


## Exercises

1. Take covariates from dataset XYZ and simulate data with 5 active covariates. Run forward regression, 
ridge regression and lasso regression

2. Calculate the Ridge and the Lasso solution for orthogonal covariates.

3. 
