---
title: "Exercises"
author: "Nicolas Städler"
date: "3/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE)
```


```{r, echo=FALSE}
library(knitr)
library(caret)
library(glmnet)
library(tidyverse)
library(MASS)
library(lars)
library(gridExtra)
```



## Another Example

We simulate data for training and testing.

```{r}
set.seed(1)
n <- 20
p <- 50
# simulate covariates
xtrain <- matrix(rnorm(n*p),n,p)
xtest <- matrix(rnorm(n*p),n,p)
# simulate outcome
beta0 <- 0
beta <- c(2,0,0,0,2,rep(0,p-5))
ytrain <- beta0+xtrain%*%beta+rnorm(n,sd=0.5)
ytest <- beta0+xtest%*%beta+rnorm(n,sd=0.5)
# train and test data as data.frames
dtrain <- data.frame(cbind(ytrain,xtrain))
colnames(dtrain) <- c("y",paste0("x",1:ncol(xtrain)))
dtest <- data.frame(cbind(ytest,xtest))
colnames(dtest) <- c("y",paste0("x",1:ncol(xtrain)))
```

Linear regression with only x1.

```{r}
fit <- lm(y~x1,data=dtrain)
t.gp1 <- dtrain%>%
  ggplot(.,aes(x=x1))+
  geom_point(aes(y=y))+
  geom_line(aes(y=predict(fit)))
t.gp2 <- dtest%>%
  ggplot(aes(x=x1,y=y))+
  geom_point()+
  geom_line(aes(x=x1,y=predict(fit,newdata = dtest)),col="red")
grid.arrange(t.gp1,t.gp2,ncol=2)
```


Linear regression with 10 covariates.
```{r}
fm <- as.formula(paste0("y~",paste0("x",1:10,collapse="+")))
fit <- lm(fm,data=dtrain)
t.gp1 <- dtrain%>%
  ggplot(.,aes(x=x1))+
  geom_point(aes(y=y))+
  geom_line(aes(y=predict(fit)))
t.gp2 <- dtest%>%
  ggplot(aes(x=x1,y=y))+
  geom_point()+
  geom_line(aes(x=x1,y=predict(fit,newdata = dtest)),col="red")
grid.arrange(t.gp1,t.gp2,ncol=2)
```

Linear regression with 19 covariates
```{r}
fm <- as.formula(paste0("y~",paste0("x",1:19,collapse="+")))
fit <- lm(fm,data=dtrain)
t.gp1 <- dtrain%>%
  ggplot(.,aes(x=x1))+
  geom_point(aes(y=y))+
  geom_line(aes(y=predict(fit)))
t.gp2 <- dtest%>%
  ggplot(aes(x=x1,y=y))+
  geom_point()+
  geom_line(aes(x=x1,y=predict(fit,newdata = dtest)),col="red")
grid.arrange(t.gp1,t.gp2,ncol=2)
```

Leap Forward Regression.


```{r, echo=FALSE}
tc <- trainControl(method = "cv", number = 10)
fit.forward <- train(y ~., data = dtrain,
                     method = "leapForward", 
                     tuneGrid=data.frame(nvmax=1:15),
                     trControl = tc,
                     trace = FALSE
                    )
# nvmax
fit.forward$bestTune
# accuracy
fit.forward$results
# summary of the model
summary(fit.forward$finalModel)
# final model coefficients
coef(fit.forward$finalModel,2)
```


Now we perform Ridge regession

```{r}
lambda.max <- 10
lambda.min <- 10^{-6}*lambda.max
lambda.grid <- 10^seq(log10(lambda.min),log10(lambda.max),length=100)
fit.ridge<-train(y ~., 
                 data = dtrain,
                 method = "glmnet",
                 tuneGrid = expand.grid(alpha = 0,lambda=lambda.grid),
                 trControl = tc
) 

# Cv plot
plot(fit.ridge)
# Best lambda
fit.ridge$bestTune$lambda
# Model coefficients
coef(fit.ridge$finalModel,fit.ridge$bestTune$lambda)%>%head
# Make predictions
fit.ridge %>% predict(dtest)%>%head
```


```{r}
lambda.max <- max(abs(t(xtrain)%*%ytrain))/nrow(xtrain)
lambda.min <- 10^{-6}*lambda.max
lambda.grid <- 10^seq(log10(lambda.min),log10(lambda.max),length=100)
fit.lasso<-train(y ~., 
                 data = dtrain,
                 method = "glmnet",
                 tuneGrid = expand.grid(alpha = 1,lambda=lambda.grid),
                 trControl = tc
) 

# Cv plot
plot(fit.lasso)
# Best lambda
fit.lasso$bestTune$lambda
# Model coefficients
coef(fit.lasso$finalModel,fit.lasso$bestTune$lambda)%>%head
# Make predictions
fit.lasso %>% predict(dtest)%>%head
```


```{r}
models <- list(forward= fit.forward,ridge = fit.ridge,lasso=fit.lasso)
resamples(models) %>% summary( metric = "RMSE")
```

That is how to calculate lambda_max
```{r}
# with scaling
mysd <- function(y) sqrt(sum((y-mean(y))^2)/length(y))
sx <- scale(xtrain,scale=apply(xtrain, 2, mysd))
sx <- as.matrix(sx, ncol=20, nrow=100)
sy <- as.vector(scale(ytrain, scale=mysd(ytrain)))
max(abs(colSums(sx*sy)))/nrow(xtrain)
fitglmnet <- glmnet(sx,sy)
max(fitglmnet$lambda)

# no scaling
fit.lasso <-glmnet(x=xtrain,y=ytrain,
                   alpha=1,
                   standardize=FALSE,
                   intercept=FALSE) 
max(fit.lasso$lambda)
max(abs(t(xtrain)%*%ytrain))/nrow(xtrain)
```

### Example based on simulated data

Take covariates from dataset XYZ and simulate data with 5 active covariates. Run forward regression, 
ridge regression and lasso regression

### Riboflavin data set from hdi package

### Ridge and Lasso for the orthonormal design

Calculate the Ridge and the Lasso solution for orthogonal covariates.

### Log-likelihood of Regression Model

Write down the log-likelihood of the model
$Y_i=X_{i}^T\beta +\epsilon_i,$ where $\epsilon_1,\epsilon_n$ iid $N(0,\sigma^2)$ and $\bf{X}$ is a fixed
$n\times p$ design matrix.

### P-values for High-dimensional Regression

